\thetitle

### 2021-02-12 writing/landscape.tex
touching up the prose

Under the Natural semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017}, when
\texttt{pack} is imported in either \texttt{typed/pack-lib} or
\texttt{typed/secure-pack-lib}, the runtime checks that \texttt{pack} is a
function and wraps it in a proxy that enforces its  type.  Analogously
the runtime wraps the exports of the two typed modules in proxies that
check that untyped code applies \texttt{typed-pack} and
\texttt{crypto-pack} to arguments of the expected type. In essence, the
return type checks protect  the typed modules in case they use
\texttt{pack} and, as is the case here, it does not return an association
list. The check of the argument type  treats the
exports of the two typed modules as typed values that need to be protected
when they are used by untyped code.  As this analysis implies, if the
return type check fails then the problem is with
\texttt{untyped/pack-lib}, while if the argument type check fails
responsibility lies with \texttt{client}. The Natural semantics keep track
of the responsible party for each check in the corresponding proxy. Thus,
in our example, when the call to \texttt{typed-pack} returns, the return
type check fails and the Natural semantics blames \texttt{pack-lib}.

Under the Transient semantics~\cite{vss-popl-2017}, the call to \texttt{typed-pack} does not
produce an error. In contrast to Natural, Transient does not rely on
proxies. Instead, it rewrites typed modules to inline checks at appropriate
places. Furthermore, these checks are
partial, that is, they check only the top constructor of a type. In other words
they check types in a  \emph{shallow} manner. In our
example, Transient's inlined checks make sure that the imported \texttt{pack} 
is a function and that the result of its call in \texttt{typed/crypto-pack-lib} is a list.   Exactly this
last check fails when \texttt{client} calls to \texttt{crypto-pack}. In
order to pinpoint the responsible party for the failed check, Transient
keeps on the side a map from values to the boundaries between modules they
cross and the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{untyped/pack-lib} to
\texttt{typed/pack-lib} and from \texttt{typed/pack-lib} to
\texttt{client} with the type that appears in the \texttt{required/typed}
forms in the example. Since the failed check corresponds to the return
type of \texttt{pack}, Transient blames the sources of the two boundary
crosses, which in this case is \texttt{untyped/pack-lib} for both. In
general though, Transient may blame more than one party. In fact, the
theoretical work of~\citet{gfd-oopsla-2019} shows that there are programs
where Transient constructs a blame set that includes modules that are not
relevant to the check failure or that it even misses entirely the responsible
party ---  our experiment aims to examine the pragmatic impact of exactly these
issues on debugging type-level mistakes.

Under the Erasure semantics, the example does not raise an error due to the
gradual type system; Erasure performs no type checks at runtime. Thus the
program may fail with an exception if \texttt{client} tries to use the
result of \texttt{typed-pack} or \texttt{crypto-pack}, but it may also not
fail at all and produce a wrong result that goes undetected by the
gradual type system .  In either case, the Erasure semantics provides no
help to the programmer for detecting the bug in our program.

@@

The first major ingredient of Transient Typed Racket is to determine the
Racket implementation of the shallow checks that map to each type.  For
certain types, such as {\tt String} or {\tt (Vectorof Byte)}, Transient
can re-use a predicate from the Racket language.  Other types require custom
predicates; for example the type {\tt Index} of valid array indices allows
a range of positive exact integers.  Many custom predicates run in
constant time, but some are more involved.  One linear-time check is for
{\tt (Listof T)} for any element type {\tt T}, because it must distinguish
proper from improper list-like structures.  The
key point, however, is that no shallow check inspects beyond the top
constructor of each type.

The second major ingredient of Transient Typed Racket  is the addition of
a pass to the compiler of Typed Racket that rewrites all typed code.
Rewriting modifies all elimination forms to shallow-check their result.
For example, if the variable {\tt f} has the static type {\tt (-> String
String)} then the application {\tt (f x)} must be rewritten to check for a
string result because {\tt f} could be an untyped function.  Rewriting
also inserts hooks to update the blame map.  When a value crosses a module
boundary, the blame map records a type assumption.  The inlined checks
for elimination forms also extend the blame map with further information. For
instance, the check of the result of {\tt (f x)} adds to the blame map
that the expected type of the result is the return type of the type of
{\tt f}.  If a Transient check ever fails, Transient Typed Racket walks over 
the information in the blame map and reports a set of blamed components.

\smallskip
\noindent
{\bf Note.} As the common basis for the three semantics, we choose 
Typed Racket's migratory typing. In contrast to the initial formulation of
gradual typing~\cite{st-sfp-2006}, migratory typing does not provide
support for type \texttt{Dyn}, components are either untyped or fully
typed and the gradual type system does not affect the compilation of
untyped components. Despite, these differences, Typed Racket is the most
mature system for sound mix-typed code
and recent theoretical work has established that
its migratory typing offers a setting for comparing different approaches to gradual
typing~\cite{gf-icfp-2018,gfd-oopsla-2019}.

>>>>

Under the Natural semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017}, when
\texttt{pack} is imported in either \texttt{typed/pack-lib} or
\texttt{typed/secure-pack-lib}, the runtime checks that \texttt{pack} is a
function and wraps it in a proxy that enforces its  type.  Analogously
the runtime wraps the exports of the two typed modules in proxies that
check that untyped code applies \texttt{typed-pack} and
\texttt{crypto-pack} to arguments of the expected type. In essence, the
return type checks protect  the typed modules in case they use
\texttt{pack} and, as is the case here, it does not return an association
list. The check of the argument type  treats the
exports of the two typed modules as typed values that need to be protected
when they are used by untyped code.  As this analysis implies, if the
return type check fails then the problem is with
\texttt{untyped/pack-lib}, while if the argument type check fails
responsibility lies with \texttt{client}. The Natural semantics keep track
of the responsible party for each check in the corresponding proxy. Thus,
in our example, as the call to \texttt{typed-pack} returns, the return
type check fails and the Natural semantics blames \texttt{pack-lib}.

Under the Transient semantics~\cite{vss-popl-2017}, the call to \texttt{typed-pack} does not
produce an error. In contrast to Natural, Transient does not rely on
proxies. Instead, it rewrites typed modules to inline checks at appropriate
places. Furthermore, these checks are
partial, that is, they check only the top constructor of a type. In other words
they check types in a  \emph{shallow} manner. In our
example, Transient's inlined checks make sure that the imported \texttt{pack} 
is a function and that the result of its call in \texttt{typed/crypto-pack-lib} is a list.   Exactly this
last check fails when \texttt{client} calls to \texttt{crypto-pack}. In
order to pinpoint the responsible party for the failed check, Transient
keeps on the side a map from values to the boundaries between modules they
cross and the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{untyped/pack-lib} to
\texttt{typed/pack-lib} and from \texttt{typed/pack-lib} to
\texttt{client} with the type that appears in the \texttt{required/typed}
forms in the example. Since the failed check corresponds to the return
type of \texttt{pack}, Transient blames the sources of the two
boundary crosses, which in this case is \texttt{untyped/pack-lib} for
both. In general though, Transient may blame more than one party. In
fact, the theoretical work of~\citet{gfd-oopsla-2019} shows that there
are programs where Transient constructs a blame set that includes
modules that are not relevant to the check failure or that it even
entirely misses the responsible party--our experiment aims to examine
the pragmatic impact of exactly these issues on debugging type-level
mistakes.

Under the Erasure semantics, the example does not raise an error due
to the gradual type system; Erasure performs no type checks at
runtime. Thus the program may fail with an exception if
\texttt{client} tries to use the result of \texttt{typed-pack} or
\texttt{crypto-pack}, but it may also not fail at all and produce a
result of the proper kind---and yet is completely wrong.  In either
case, the Erasure semantics provides no help to the programmer for
detecting the bug in our program.

@@

The first major ingredient to the construction of Transient Typed Racket is to determine the
Racket implementation of the shallow checks that map to each type.  For
certain types, such as {\tt String} or {\tt (Vectorof Byte)}, Transient
can re-use a predicate from the Racket language.  Other types require custom
predicates; for example the type {\tt Index} of valid array indices allows
a range of positive exact integers.  Many custom predicates run in
constant time, but some are more involved.  One linear-time check is for
{\tt (Listof T)} for any element type {\tt T}, because it must distinguish
proper from improper list-like structures.  The
key point, however, is that no shallow check inspects beyond the top
constructor of each type.

The second major ingredient of Transient Typed Racket  is the addition of
a pass to the compiler of Typed Racket that rewrites all typed code.
Rewriting modifies all elimination forms to shallow-check their result.
For example, if the variable {\tt f} has the static type {\tt (-> String
String)} then the application {\tt (f x)} must be rewritten to check for a
string result because {\tt f} could be an untyped function.  Rewriting
also inserts hooks to update the blame map.  When a value crosses a module
boundary, the blame map records a type assumption.  The inlined checks
for elimination forms also extend the blame map with further information. For
instance, the check of the result of {\tt (f x)} adds an entry to the blame map
that records the expected type of the result as the return type of the type of
{\tt f}.  If a Transient check ever fails, Transient Typed Racket walks over 
the information in the blame map and reports a set of blamed components.

\smallskip
\noindent
{\bf Note.} As the common basis for the three semantics, we choose 
Typed Racket's migratory typing. In contrast to the initial formulation of
gradual typing~\cite{st-sfp-2006}, migratory typing does not provide
support for type \texttt{Dyn}, components are either untyped or fully
typed.
%% MF: does it ever:
%% and the gradual type system does not affect the compilation of untyped components. 
Despite, these differences, Typed Racket is the most
mature system for sound mix-typed code
and recent theoretical work has established that
its migratory typing offers a setting for comparing different approaches to gradual
typing~\cite{gf-icfp-2018,gfd-oopsla-2019}.

### 2021-02-12 writing/mutate.tex
Introduce transient in intro, reword "mere typos"
touching up the prose

An important challenge from section~\ref{sec:challenges} is the generation of
mix-typed programs with representative impedance mismatches between
the typed components, their types, and the behavior of untyped components. In this section, we
present a generation process and argue that, despite its synthetic
nature, it yields a large corpus of programs with  realistic and diverse type mistakes.

@@

Of course, the gradual typing benchmarks have no mistakes for the rational
programmer to debug. Therefore, we follow ~\citet{lksfd-popl-2020} and we
inject bugs with mutation
analysis~\cite{jia2011analysis,demillo1978hints,lipton1971fault}.  A
mutation is a local syntactic change of the code of a program that may
produce a bug. Usually, the outcome of mutating a program once, dubbed a
mutant, can be distinguished from the original program with a test case.
The better the test suite of a program, the more mutants it can ``kill.''
Mutants that are hard to ``kill'' usually correspond to subtle programming
errors, and mutation frameworks typically come with standard mutators
carefully designed to create such mutants.

However,  unlike the hard-to-kill mistakes introduced by standard mutators, 
gradual type systems flag mere typos.  Thus, in order to
exercise the rational programmer in the context of gradual typing, 
we need to develop alternative mutators
that are fine-tuned to produce the type-level mistakes gradual type systems 
can detect. Our inspiration is our
more-than-a-decade long experience of making type-level mistakes in Typed
Racket. Some of these mistakes often take non-trivial effort to debug and
have become the seed for our mutators, which we summarize in
figure~\ref{table:mutation-ops}.  For each one, the figure provides a
short description and an example mutation.

@@

The astute reader must have observed that mutators that produce 
type-level mistakes do not necessarily turn our benchmarks into
interesting debugging scenarios for the rational programmer.  The
benchmarks are fully typed, and type checking detects each mistake at
compile time.  Hence, we apply the mutators to versions, dubbed
\dubbed{scenarios}, of the benchmarks without 
some  of their type annotations.  Without the annotations, the type
checker may not detect a mistake at compile time and, instead, the run
time checks
of the gradual type system signal an error.  Scenarios
with type-level mistakes that result in run time errors are the ideal
ground to exercise the rational programmer's ability to locate errors.

@@

The first criterion simply validates that the debugging scenario concerns
an actual type-level mistake. However, The second and third criteria require 
more explanation.  The second one stipulates that the mistake is
detectable under Erasure to ensure that it is detectable by all three
semantics.   Whether a scenario produces a run time error largely depends on 
the underlying semantics but \citet{gfd-oopsla-2019}
show that a run time error under Erasure implies a run time
error under both Transient and Natural. Thus the criterion aligns
interesting debugging scenarios with our lowest denominator, Erasure, 
and excludes  from consideration the very interesting scenarios indeed where Natural or
Transient can detect a mistake but where Erasure is powerless. Of course this choice favors
Erasure over Transient and Natural and, for the same reason, Transient over
Natural. We consider some form of bias towards one or the other semantics
unavoidable no matter how we define interesting debugging scenarios and we
opt for tipping the scales in favor of the theoretically weakest
semantics. We discuss further the implications of this choice in
section~\ref{sec:discussion}.

>>>>

%% this is awkward, but what I removed was exponentially awkward

In this section, we
present a generation process and argue that, despite its synthetic
nature, it yields a large corpus of programs with  realistic and diverse type mistakes.

@@

Of course, the gradual typing benchmarks have no mistakes for the rational
programmer to debug. Therefore, we follow ~\citet{lksfd-popl-2020} and we
inject bugs with mutation
analysis~\cite{jia2011analysis,demillo1978hints,lipton1971fault}.  A
mutation is a local syntactic change of the code that may
produce a bug. Usually, the outcome of mutating a program once, dubbed a
mutant, can be distinguished from the original program with a test case.
The better the test suite of a program, the more mutants it can ``kill.''
Mutants that are hard to ``kill'' usually correspond to subtle programming
errors, and mutation frameworks typically come with standard mutators
carefully designed to create such mutants.

Unlike the hard-to-kill mistakes introduced by standard mutators, 
gradual type systems catch shallow errors.  Thus, in order to
exercise the rational programmer in the context of gradual typing, 
we need to develop alternative mutators
that are fine-tuned to produce the type-level mistakes gradual type systems 
can detect. Our inspiration is our
more-than-a-decade long experience of making type-level mistakes in Typed
Racket. Some of these mistakes often take non-trivial effort to debug and
have become the seed for our mutators, which we summarize in
figure~\ref{table:mutation-ops}.  For each one, the figure provides a
short description and an example mutation.

@@

The reader may have observed that mutators that produce 
type-level mistakes do not necessarily turn our benchmarks into
interesting debugging scenarios for the rational programmer.  The
benchmarks are fully typed, and type checking detects each mistake at
compile time.  Hence, we apply the mutators to versions, dubbed
dubbed{scenarios}, of the benchmarks without 
some  of their type annotations.  Without the annotations, the type
checker may not detect a mistake at compile time and, instead, the run
time checks
of the gradual type system signal an error.  Scenarios
with type-level mistakes that result in run time errors are the ideal
ground to exercise the rational programmer's ability to locate errors.

@@

The first criterion simply validates that the debugging scenario concerns
an actual type-level mistake. The second one stipulates that the mistake is
detectable under Erasure to ensure that it is detectable by all three
semantics.   Whether a scenario produces a run time error largely depends on 
the underlying semantics but \citet{gfd-oopsla-2019}
show that a run time error under Erasure implies a run time
error under both Transient and Natural. Thus the criterion aligns
interesting debugging scenarios with our lowest denominator, Erasure, 
and excludes  from consideration the very interesting scenarios indeed where Natural or
Transient can detect a mistake but where Erasure is powerless. Of course this choice favors
Erasure over Transient and Natural and, for the same reason, Transient over
Natural. We consider some form of bias towards one or the other semantics
unavoidable no matter how we define interesting debugging scenarios and we
opt for tipping the scales in favor of the theoretically weakest
semantics. We further discuss the implications of this choice in
section~\ref{sec:discussion}.
